Improving disaggregation models of malaria incidence by ensembling non-linear models of prevalence

### 2020-01-09 tim-lucas-model-ensembles.tex
Cite WMR.

Maps of infection rate are one of the core tools needed for the control and elimination of malaria.
Reliable routine surveillance data of malaria incidence is becoming more widely available.
Disaggregation regression is becoming an important model framework for estimating high resolution risk maps from these aggregated data.
However, the aggregation of incidence over large, heterogeneous areas means that these data are underpowered for learning complex, non-linear relationships between the environment and malaria risk.
In contrast, prevalence point-surveys are directly linked to local environmental conditions but are not common in many areas of the world.
A model that combines point surveys and aggregated surveillance data could have the benefits of both but must be able to account for the fact that these two data types measure different malariometric units.
Here, we train multiple non-linear, machine learning models on \emph{Plasmodium falciparum} prevalence point-surveys.
We then ensemble the predictions from these machine learning models with a geostatistical disaggregation regression model that uses malaria incidence, aggregated to administrative units, as response data.
We find that, in four national case studies, using a disaggregation regression model to combine predictions from machine learning models trained on prevalence point-surveys improves model accuracy relative to a baseline disaggregation regression model that only uses linear effects of the raw environmental covariates.

@@

High-resolution maps of malaria risk are vital for control and elimination \citep{weiss2019mapping, battle2019mapping}.
However, mapping malaria in lower burden countries presents new challenges as traditional mapping of prevalence from cluster-level surveys \cite{weiss2019mapping,battle2019mapping, bhatt2017improved, bhatt2015effect} is often not effective for two reasons.
Firstly, so few individuals are infected that most surveys will detect zero cases \citep{sturrock2016mapping}.
Secondly, there is a lack of nationally representative prevalence surveys in low burden countries \citep{sturrock2016mapping, sturrock2014fine}. 
Routine surveillance data of malaria case counts, often aggregated over administrative regions defined by geographic polygons, is becoming more reliable and more widely available \citep{sturrock2016mapping} and recent work has focussed on methods for estimating high-resolution malaria risk from these data \citep{sturrock2014fine, wilson2017pointless, law2018variational, taylor2017continuous, li2012log, johnson2019spatially}. 
However, the aggregation of cases over space means that the data may be spatially uninformative, especially if the case counts are aggregated over large or heterogeneous areas, because it is unclear where within the polygon, and in which environments, the cases occurred. 
This data is therefore often under-powered for fitting flexible, non-linear models as is required for accurate malaria maps \citep{bhatt2017improved, bhatt2015effect}. 
A method that combines prevalence point-surveys and aggregated surveillance data, and therefore leverages the strength of both, has great potential.

@@

To test the effectiveness of the proposed approach we used data from four countries: Madagascar, Colombia, Indonesia and Senegal.
We focused our analysis on comparing the predictive performance of disaggregation regression when given different sets of covariates.
Therefore we keep the structure of the disaggregation regression model the same and only vary the covariates provided to the model.
In each country we fitted stage 1 machine learning models trained on prevalence data and raw environmental covariates.
We made new covariates using predictions from these models.
We then tested whether stage 2 disaggregation regression models with these new covariates performed better than a baseline disaggregation regression model that directly used the raw environmental covariates.
We tested this approach using machine learning models trained on local prevalence data as well as models trained on a global prevalence dataset.
While there was no consistently best model we found that, in most cases, the two stage method worked better than the single stage baseline disaggregation regression models.
Using predictions from machine learning models, trained on local prevalence data as covariates improved the performance of disaggregation regression models relative to the disaggregation regression models that only used the raw environmental covariates.
In contrast, using predictions from machine learning models trained on the global prevalence dataset rarely improved predictive performance.

@@

The polygon incidence data (i.e. malaria incidence aggregated to administrative units) were collected from government reports \citep{mdg, sen, col, idn} and standardised using methods defined in \cite{cibulskis2011worldwide}.
This standardisation step accounts for missed cases due to lack of treatment seeking, missing case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}.
For reports where cases were not reported at the species level, national estimates of the ratio between \emph{P. falciparum} and \emph{Plasmodium vivax} cases were used to calculate \emph{P. falciparum} only cases. 
For incidence rates we divide by 1000 to give the Annual Parasite Index (API).
To keep the analysis focused on spatial estimates we selected one year of surveillance data for each country. 
We used annual surveillance data from 2013 for Madagascar (110 districts), 2015 for Colombia (952 municipalities), 2013 for Indonesia (244 regencies and cities) and 2009 for Senegal (34 departments).
These years were selected as they had the most complete data in each case.

@@

The polygon case count data, $y_a$ is given a Poisson likelihood
$$y_a \sim \operatorname{Pois}(i_a\mathrm{pop_a})$$
where $i_a$ is the estimated polygon incidence rate and $\mathrm{pop_a}$ is the observed polygon population-at-risk. 
This polygon-level likelihood is linked to the pixel-level incidence and prevalence by
$$i_a = \frac{ \sum(i_b \mathrm{pop}_b)}{\sum  \mathrm{pop}_b} $$
$$i_b = 2.616p_b - 3.596p_b^2 + 1.594p_b^3$$
where the polynomial is  a function from a previously published model \citep{cameron2015defining}.
The fact that the model passes through prevalence space ensures that the predictions from the machine learning models can be appropriately scaled.
The linear predictor of the model is related to prevalence by a typical logit link function and includes: an intercept, $\beta_0$; covariates, $X$, with a vector of regression parameters, $\beta$; a spatial, Gaussian, random field, $u_b(\rho, \sigma_u)$; and an \emph{iid} random effect, $v_a(\sigma_v)$.
$$p_b = \operatorname{logit}^{-1}\left(\beta_0 + \beta X  + u_b\left(\rho, \sigma_u\right) + v_a\left(\sigma_v\right)\right)$$
The Gaussian spatial effect has a Mat\'ern covariance function and two hyperparameters: $\rho$, the nominal range (beyond which correlation is $< 0.1$) and $\sigma_u$, the marginal standard deviation.
The \emph{iid} random effect models both unobserved explanatory factors and extra-Poisson sampling error.
As described in more detail below we do not vary the structure of this model in the methodological comparison.
The matrix of covariates, $X$, it the only component that varies and is made up of various combinations of raw environmental covariates and predictions from the machine learning models.

@@

Table \ref{t:results} gives the correlation between observed and held out data (under random and spatial cross-validation) for experiment \emph{i} (models Enviro, ML\textsubscript{l} and Envir + ML\textsubscript{l}).
Many of the differences in performance are rather marginal.
Enviro was the best performing model in two cases (random cross-validation in Madagascar and Senegal).
In one case, Enviro perfomed equally well as another model; under spatial cross-validation in Senegal Enviro and Enviro + ML\textsubscript{l} performed equally well.
Of the remaining five cases, in two cases ML\textsubscript{l} performed best and in three cases Envir + ML\textsubscript{l} performed best.
The greatest benefits to using prediction from machine learning models instead of, or in combination with, the raw environmental variables occured under spatial cross-validation and in the cases when the Enviro model was particularly poor.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{figs/cv1_l_scatter.png}
\caption{
  Observed data against predictions for random cross-validation hold-out samples on a square root transformed scale. 
  There are 12 cases composed of 4 countries (COL:, Colombia, IDN: Indonesia, MDG: Madagascar, SEN: Senegal) and three sets of covariates (Envir: raw environmental covariates only, Envir +  ML\textsubscript{l}: raw environmental covariates and machine learning covariates trained on local prevalence data combined, ML\textsubscript{l}: Machine learning models trained on local prevalence data only.
}
\label{f:scatter1}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{figs/cv2_l_scatter.png}
\caption{
  Observed data against predictions for spatial cross-validation hold-out samples on a square root transformed scale.
  There are 12 cases composed of 4 countries (COL:, Colombia, IDN: Indonesia, MDG: Madagascar, SEN: Senegal) and three sets of covariates (Envir: raw environmental covariates only, Envir +  ML\textsubscript{l}: raw environmental covariates and machine learning covariates trained on local prevalence data combined, ML\textsubscript{l}: Machine learning models trained on local prevalence data only.
}
\label{f:scatter2}
\end{figure}

@@

Table~\ref{t:coverage} shows the out-of-sample coverage of the 80\% credible intervals for all models, countries and cross-validation schemes.
The coverage in Colombia was very poor with no models achieving a coverage above 0.4.
The coverage was acceptable in the other three countries with most values lying between 0.7 and 0.9.
Overall there was a general tendancy for models to be slightly over confident.

@@

Overall, the models fitted using predictions from machine learning models trained on the global database of prevalence point-surveys were less good than those using either environmental data alone or than those using predictions from machine learning models trained on local data.
This in itself is not particularly surprising.
However, it does indicate that continental scale models should consider using a mosaic of locally trained machine learning models for example.
What is more surprising is that using both the ML\textsubscript{l} and ML\textsubscript{g} covariates together did not improve performance.
Given that this approach of creating diverse predictions is a core element of stacking methodology we would expect this set of covariates to perform as well or better than just the ML\textsubscript{l} covariates.

>>>>

Maps of disease burden are one of the core tools needed for the control and elimination of malaria.
Reliable routine surveillance data of malaria incidence, typically aggregated to administrative units, is becoming more widely available.
Disaggregation regression is becoming an important model framework for estimating high resolution risk maps from these aggregated data.
However, the aggregation of incidence over large, heterogeneous areas means that these data are underpowered for learning complex, non-linear relationships between the environment and malaria risk.
In contrast, prevalence point-surveys are directly linked to local environmental conditions but are not common in many areas of the world.
A model that combines point surveys and aggregated surveillance data could have the benefits of both but must be able to account for the fact that these two data types measure different malariometric units.
Here, we train multiple non-linear, machine learning models on \emph{Plasmodium falciparum} prevalence point-surveys.
We then ensemble the predictions from these machine learning models with a geostatistical disaggregation regression model that uses malaria incidence, aggregated to administrative units, as response data.
We find that, in four national case studies, using a disaggregation regression model to combine predictions from machine learning models trained on prevalence point-surveys improves model accuracy relative to a baseline disaggregation regression model that only uses linear effects of the raw environmental covariates.

@@

High-resolution maps of malaria risk are vital for control and elimination \citep{weiss2019mapping, battle2019mapping}.
However, mapping malaria in lower burden countries presents new challenges as traditional mapping of prevalence from cluster-level surveys \cite{weiss2019mapping,battle2019mapping, bhatt2017improved, bhatt2015effect} is often not effective for two reasons.
Firstly, so few individuals are infected that most surveys will detect zero positives \citep{sturrock2016mapping}.
Secondly, there is a lack of nationally representative prevalence surveys in low burden countries \citep{sturrock2016mapping, sturrock2014fine}. 
Routine surveillance data of malaria case counts, often aggregated over administrative regions defined by geographic polygons, is becoming more reliable and more widely available \citep{sturrock2016mapping} and recent work has focussed on methods for estimating high-resolution malaria risk from these data \citep{sturrock2014fine, wilson2017pointless, law2018variational, taylor2017continuous, li2012log, johnson2019spatially}. 
However, the aggregation of cases over space means that the data may be spatially uninformative, especially if the case counts are aggregated over large or heterogeneous areas, because it is unclear where within the polygon, and in which environments, the cases occurred. 
This data is therefore often under-powered for fitting flexible, non-linear models as is required for accurate malaria maps \citep{bhatt2017improved, bhatt2015effect}. 
A method that combines prevalence point-surveys and aggregated surveillance data, and therefore leverages the strength of both, has great potential.

@@

To test the effectiveness of the proposed approach we used data from four countries with relatively complete surveillance data: Madagascar, Colombia, Indonesia and Senegal.
We focused our analysis on comparing the predictive performance of disaggregation regression when given different sets of covariates.
Therefore we keep the structure of the disaggregation regression model the same and only vary the covariates provided to the model.
In each country we fitted stage 1 machine learning models trained on prevalence data and raw environmental covariates.
We made new covariates using predictions from these models.
We then tested whether stage 2 disaggregation regression models with these new covariates performed better than a baseline disaggregation regression model that directly used the raw environmental covariates.
We tested this approach using machine learning models trained on local prevalence data as well as models trained on a global prevalence dataset.
While there was no consistently best model we found that, in most cases, the two stage method worked better than the single stage baseline disaggregation regression models.
Using predictions from machine learning models, trained on local prevalence data as covariates improved the performance of disaggregation regression models relative to the disaggregation regression models that only used the raw environmental covariates.
In contrast, using predictions from machine learning models trained on the global prevalence dataset rarely improved predictive performance.

@@

The polygon incidence data (i.e. malaria incidence aggregated to administrative units) were collected from government reports \citep{mdg, sen, col, idn} and standardised using methods defined in \cite{cibulskis2011worldwide}.
This standardisation step accounts for missed cases due to lack of treatment seeking, missing case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}.
For reports where cases were not reported at the species level, national estimates of the ratio between \emph{P. falciparum} and \emph{Plasmodium vivax} cases from the World Malaria Report were used to calculate \emph{P. falciparum} only cases \citep{world2016world}. 
For incidence rates we divide by 1000 to give the Annual Parasite Index (API).
To keep the analysis focused on spatial estimates we selected one year of surveillance data for each country. 
We used annual surveillance data from 2013 for Madagascar (110 districts), 2015 for Colombia (952 municipalities), 2013 for Indonesia (244 regencies and cities) and 2009 for Senegal (34 departments).
These years were selected as they had the most complete data in each case.

@@

The polygon case count data, $y_a$ is given a Poisson likelihood
$$y_a \sim \operatorname{Pois}(i_a\mathrm{pop_a})$$
where $i_a$ is the estimated polygon incidence rate and $\mathrm{pop_a}$ is the observed polygon population-at-risk. 
This polygon-level likelihood is linked to the pixel-level incidence and prevalence by
$$i_a = \frac{ \sum(i_b \mathrm{pop}_b)}{\sum  \mathrm{pop}_b} $$
$$i_b = 2.616p_b - 3.596p_b^2 + 1.594p_b^3$$
where the polynomial is  a function from a previously published model \citep{cameron2015defining}.
The fact that the model is explicit about the relationship between prevalence and incidence has two advantages.
Firstly, predictions of prevalence can be easily made directly from the linear predictor of the model.
Secondly, it means that the logit-transformed predictions from the machine learning models are correctly scaled.
The linear predictor of the model is related to prevalence by a typical logit link function and includes: an intercept, $\beta_0$; covariates, $X$, with a vector of regression parameters, $\beta$; a spatial, Gaussian, random field, $u_b(\rho, \sigma_u)$; and an \emph{iid} random effect, $v_a(\sigma_v)$.
$$p_b = \operatorname{logit}^{-1}\left(\beta_0 + \beta X  + u_b\left(\rho, \sigma_u\right) + v_a\left(\sigma_v\right)\right)$$
The Gaussian spatial effect has a Mat\'ern covariance function and two hyperparameters: $\rho$, the nominal range (beyond which correlation is $< 0.1$) and $\sigma_u$, the marginal standard deviation.
The \emph{iid} random effect models both unobserved explanatory factors and extra-Poisson sampling error.
As described in more detail below we do not vary the structure of this model in the methodological comparison.
The matrix of covariates, $X$, it the only component that varies and is made up of various combinations of raw environmental covariates and predictions from the machine learning models.

@@

Table \ref{t:results} gives the correlation between observed and held out data (under random and spatial cross-validation) for experiment \emph{i} (models Enviro, ML\textsubscript{l} and Enviro + ML\textsubscript{l}).
Many of the differences in performance are rather marginal.
Enviro was the best performing model in two cases (random cross-validation in Madagascar and Senegal).
In one case, Enviro perfomed equally well as another model; under spatial cross-validation in Senegal Enviro and Enviro + ML\textsubscript{l} performed equally well.
Of the remaining five cases, in two cases ML\textsubscript{l} performed best and in three cases Enviro + ML\textsubscript{l} performed best.
The greatest benefits to using prediction from machine learning models instead of, or in combination with, the raw environmental variables occured under spatial cross-validation and in the cases when the Enviro model was particularly poor.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{figs/cv1_l_scatter.png}
\caption{
  Observed data against predictions for random cross-validation hold-out samples on a square root transformed scale. 
  There are 12 cases composed of 4 countries (COL:, Colombia, IDN: Indonesia, MDG: Madagascar, SEN: Senegal) and three sets of covariates (Envir: raw environmental covariates only, Enviro +  ML\textsubscript{l}: raw environmental covariates and machine learning covariates trained on local prevalence data combined, ML\textsubscript{l}: Machine learning models trained on local prevalence data only.
}
\label{f:scatter1}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{figs/cv2_l_scatter.png}
\caption{
  Observed data against predictions for spatial cross-validation hold-out samples on a square root transformed scale.
  There are 12 cases composed of 4 countries (COL:, Colombia, IDN: Indonesia, MDG: Madagascar, SEN: Senegal) and three sets of covariates (Envir: raw environmental covariates only, Enviro +  ML\textsubscript{l}: raw environmental covariates and machine learning covariates trained on local prevalence data combined, ML\textsubscript{l}: Machine learning models trained on local prevalence data only.
}
\label{f:scatter2}
\end{figure}

@@

Table~\ref{t:coverage} shows the out-of-sample coverage of the 80\% credible intervals for all models, countries and cross-validation schemes.
The coverage in Colombia was very poor with no models achieving a coverage above 0.4.
The coverage was acceptable in the other three countries with most values lying between 0.7 and 0.9.
Overall there was a general tendency for models to be slightly over confident.

@@

Overall, the models fitted using predictions from machine learning models trained on the global database of prevalence point-surveys were worse than those using either environmental data alone or than those using predictions from machine learning models trained on local data.
This in itself is not particularly surprising.
However, it does indicate that continental scale models should consider using a mosaic of locally trained machine learning models for example.
What is more surprising is that using both the ML\textsubscript{l} and ML\textsubscript{g} covariates together did not improve performance.
Given that this approach of creating diverse predictions is a core element of stacking methodology we would expect this set of covariates to perform as well or better than just the ML\textsubscript{l} covariates.
